---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# geoDeltaAudit

Geographic crosswalks are directional allocations, not inverses.  
`geoDeltaAudit` helps quantify how much a variable changes *purely* due to boundary translation and allocation rules (pathway dependence).

## What it does

- Defines step functions for common geographic transformations (e.g., ZCTA → ZIP → county)
- Runs an audit pipeline that reports:
  - fan-out and loss at each step
  - unmapped or duplicated units
  - Δx(VAR): change induced solely by transformation choices, holding the source constant

## Installation

```r
install.packages("remotes")
remotes::install_github("phinnphace/geoDeltaAudit")
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(geoDeltaAudit)
## basic example code
```

# geoDeltaAudit

Geographic crosswalks are directional allocations, not inverses. `geoDeltaAudit` helps you quantify how much a variable changes purely due to boundary translation + allocation rules (pathway dependence).

## What it does

- Defines step functions for common transformations (e.g., ZCTA → ZIP, ZIP → county)
- Runs an audit pipeline that reports:
  - fan-out / loss at each step
  - unmapped units
  - Δx(VAR): change induced solely by transformation choices (holding the underlying source constant)


## Install

```r
install.packages("remotes")
remotes::install_github("phinnphace/geoDeltaAudit")
