---
title: "Audit the Audit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Audit the Audit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  eval    = FALSE
)
```

## What This Vignette Is

A framework that only ever shows perturbation is not a framework — it is a result
generator. This vignette applies the same audit logic to a second crosswalk and reports
the result honestly: Δ*X* = 0.

The case is Census Tract 1016, Hennepin County, Minnesota. The crosswalk is the IPUMS
NHGIS 2010 Blocks to 2020 Census Tracts interpolation file for Minnesota
(`nhgis_blk2010_tr2020_27`). The variable is total population, ACS 2019–2023 5-year
estimates at the block group level.

The zero result is the point. It demonstrates that the math is correct, that crosswalk
fit to geography determines the outcome, and that this framework reports honestly when
the answer is zero.

---

## Crosswalk Selection and Rationale

IPUMS NHGIS recommends constructing geographic crosswalks from the lowest available
census geography to minimize interpolation error when transforming summary data between
boundary systems. The appropriate source level depends on the data instrument:

- **Decennial census (short-form):** blocks
- **Long-form 1990–2000 census:** block group parts
- **ACS 5-year estimates:** block groups
- **When lower levels are unavailable:** census tracts

The demonstration variable here is total population from ACS 5-year estimates. The
correct NHGIS pathway is therefore a block-group-level crosswalk. The file selected —
`nhgis_blk2010_tr2020_27` — maps 2010 Census blocks to 2020 Census tracts for
Minnesota, with allocation weights derived from block-level interpolation. Block group
weights are aggregated from block-level weights and normalized prior to use.

The crosswalk file can be downloaded from:
<https://www.nhgis.org/geographic-crosswalks>

A note on directionality: geographic crosswalks are not invertible. The block-to-tract
crosswalk used here allocates data from block geographies to tract geographies. It
cannot be reversed to allocate tract data back to blocks. The audit treats each
transformation as a one-way operation and records perturbation at each step
accordingly.

**Data citation:**
Jonathan Schroeder, David Van Riper, Steven Manson, Katherine Knowles, Tracy Kugler,
Finn Roberts, and Steven Ruggles. IPUMS National Historical Geographic Information
System: Version 20.0 [dataset]. Minneapolis, MN: IPUMS. 2025.
<http://doi.org/10.18128/D050.V20.0>

---

## A Note on IPUMS NHGIS

IPUMS NHGIS applies its allocation logic rigorously and documents it carefully. That
internal consistency is genuinely valuable. It is not the same as epistemic
completeness. The assumptions encoded in the NHGIS crosswalk are better labeled than
most — they are not absent. A well-documented crosswalk that is mismatched to the
analytic context will still perturb the variable.

This vignette does not argue that NHGIS resolves the problem the shellgame framework
describes. It uses NHGIS as a methodologically disciplined reference case to demonstrate
that the audit framework measures correctly in both directions.

---

## 1. Libraries

```{r libs}
library(dplyr)
library(readr)
library(tidycensus)
```

---

## 2. Configuration

```{r config}
# --- swap these values for your own target geography ---
TARGET_TRACT_GEOID <- "27053101600"  # Census Tract 1016, Hennepin County, MN
TARGET_COUNTY      <- "053"          # Hennepin County FIPS
TARGET_STATE       <- "27"           # Minnesota FIPS

# --- swap this path for your downloaded NHGIS crosswalk file ---
# Download: https://www.nhgis.org/geographic-crosswalks
# Select: 2010 Blocks -> 2020 Census Tracts, filter to your state
CROSSWALK_PATH <- "path/to/nhgis_blk2010_tr2020_[state].csv"

# Census API key — set once via tidycensus::census_api_key("YOUR_KEY", install = TRUE)
# Loads automatically from .Renviron in subsequent sessions
```

---

## 3. Load and Prepare the NHGIS Crosswalk

The NHGIS file provides block-level weights. Because ACS data are available at the
block group level rather than the block level, the crosswalk must first be aggregated
to block group. Weights are normalized by the total number of blocks in each block
group to produce a weight between 0 and 1.

```{r load-crosswalk}
crosswalk_raw <- readr::read_csv(CROSSWALK_PATH, show_col_types = FALSE)

# Extract block group GEOID from block GEOID (first 12 characters)
crosswalk_with_bg <- crosswalk_raw |>
  dplyr::mutate(
    bg2010ge = as.character(substr(as.character(blk2010ge), 1, 12)),
    tr2020ge = as.character(tr2020ge)
  )

# Total blocks per block group — denominator for normalization
bg_totals <- crosswalk_with_bg |>
  dplyr::group_by(bg2010ge) |>
  dplyr::summarise(total_blocks_in_bg = dplyr::n(), .groups = "drop")

# Aggregate to block group -> tract pairs and normalize weights
crosswalk_bg <- crosswalk_with_bg |>
  dplyr::group_by(bg2010ge, tr2020ge) |>
  dplyr::summarise(
    sum_weights       = sum(weight, na.rm = TRUE),
    n_blocks_to_tract = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::left_join(bg_totals, by = "bg2010ge") |>
  dplyr::mutate(
    wt_pop = sum_weights / total_blocks_in_bg  # normalized weight
  )
```

---

## 4. Filter to Target Tract

```{r filter-tract}
bgs_in_tract <- crosswalk_bg |>
  dplyr::filter(tr2020ge == TARGET_TRACT_GEOID) |>
  dplyr::arrange(dplyr::desc(wt_pop))
```

---

## 5. Retrieve ACS Block Group Data

```{r acs-data}
acs_bg_raw <- tidycensus::get_acs(
  geography = "block group",
  variables = "B01001_001",  # total population; swap for your variable
  state     = TARGET_STATE,
  county    = TARGET_COUNTY,
  year      = 2023,          # 2019-2023 5-year estimates
  survey    = "acs5",
  geometry  = FALSE
)

acs_bg <- acs_bg_raw |>
  dplyr::select(bg2010ge = GEOID, TOTPOP = estimate, TOTPOP_moe = moe)
```

---

## 6. Join and Apply Crosswalk Weights

```{r apply-weights}
analysis_data <- bgs_in_tract |>
  dplyr::left_join(acs_bg, by = "bg2010ge")

results <- analysis_data |>
  dplyr::filter(!is.na(TOTPOP)) |>
  dplyr::mutate(
    weighted_pop  = TOTPOP * wt_pop,
    weighted_moe  = TOTPOP_moe * wt_pop,
    pct_allocated = wt_pop * 100
  )

crosswalked_totpop <- sum(results$weighted_pop, na.rm = TRUE)
crosswalked_moe    <- sqrt(sum(results$weighted_moe^2, na.rm = TRUE))
original_bg_totpop <- sum(results$TOTPOP, na.rm = TRUE)
difference         <- crosswalked_totpop - original_bg_totpop
pct_difference     <- (difference / original_bg_totpop) * 100
```

---

## 7. Results

All three block groups in Tract 1016 have `wt_pop = 1.0`. Every block in each group
maps entirely into the tract with no fragmentation across boundaries. The variable
survived the transformation unchanged.

**Block group detail:**

| Block Group  | Weight | TOTPOP    | Weighted Pop | % Allocated |
|--------------|--------|-----------|--------------|-------------|
| 270531016001 | 1.0000 | 316       | 316          | 100.0%      |
| 270531016002 | 1.0000 | 945       | 945          | 100.0%      |
| 270531016003 | 1.0000 | 0         | 0            | 100.0%      |
| **TOTAL**    |        | **1,261** | **1,261**    |             |

**Summary:**

| Metric                  | Value                                             |
|-------------------------|---------------------------------------------------|
| Tract                   | Census Tract 1016, Hennepin County, MN            |
| ACS period              | 2019–2023                                         |
| Block groups            | 3                                                 |
| Crosswalk               | IPUMS NHGIS 2010 Blocks → 2020 Tracts (MN)       |
| Methodology             | Block-level weights aggregated to BG, normalized  |
| Original BG total       | 1,261 ± 439 (MOE)                                 |
| Crosswalked tract total | 1,261 ± 439 (MOE)                                 |
| Difference              | 0                                                 |
| Δ*X*(population)        | **0.00%**                                         |

---

## 8. What the Zero Result Means

Δ*X* = 0 means the variable survived this transformation unchanged. It does not mean:

- The crosswalk is correct for all use cases
- The allocation assumptions have been externally validated
- This crosswalk is appropriate for non-population variables — a crosswalk calibrated
  around block-level population geometry encodes assumptions that may not transfer to
  income, employment, housing, or other analytic contexts
- Zero perturbation is the target or the ideal

What the zero result does establish:

- The audit math is correct — when there is nothing to measure, it returns zero
- Crosswalk fit to geography determines the outcome; Tract 1016 has minimal boundary
  fragmentation and the NHGIS weights reflect that geometry cleanly
- The framework is not designed to find perturbation — it is designed to measure
  whatever is there and report it honestly

A non-zero result from a different crosswalk or a different geography is not a failure
of either crosswalk. It is a measurement. The question the audit answers is not "which
crosswalk is right" — it is "what does this crosswalk do to this variable in this
geography."

---

## What This Demonstrates

The same framework, a different crosswalk, a well-matched geography: Δ*X* = 0.

The Hennepin/HUD case returned −12.6%. This case returns 0.0%. Both results are
correct. The difference is a property of the crosswalk's fit to the geography and the
allocation assumptions it encodes — not a property of the framework, the variable, or
the analyst running it.

For the conceptual argument behind what these results mean, see
`vignette("conceptual_framework", package = "shellgame")`.

For the Hennepin County demonstration, see
`vignette("audit_walkthrough", package = "geoDeltaAudit")`.
